### 2020-08-20 latex/vlp.tex
Added the dynamics of the VLP
I managed to explain why we can ignore the dynamics of l, and treat it
as a control input instead of a configuration variable.

%/========== Variable Length Pendulum ==========/%
\chapter{Application of VNHCS: The Variable Length Pendulum}\label{sec:vlp}
\section{Motivation}
% TODO: Why is this a good problem to study? What is the problem? What would a human do on a standing swing?
\section{The VLP Constraint}
% TODO: Go through the development of the constraint and prove it gains energy

>>>>

%/========== Variable Length Pendulum ==========/%
\chapter{Application of VNHCS: The Variable Length Pendulum}\label{sec:vlp}
\section{Motivation}
The variable length pendulum (VLP) is a classical dynamical system, which is often
used to model a the motion of a person on a standing swing
\cite{dynamics_periodic_vlp,pumping_swing_standing_squatting,how_to_pump_a_swing}.
% TODO: Why else is it useful? Apparently it models cranes, satellites, etc
% TODO: Motivate more why this is a good problem to study

The variable length pendulum (VLP) is modelled as a point with mass \(m\)
connected to a pivot by a massless rod of varying length. Label
\(q \in \mathbb{S}^1\) as the angle from the vertical line to the rod. In a
realizable VLP, the rod  has length \(l\) which varies between some minimum
length \(\underline{l} \geq 0\) and some maximum length 
\(\overline{l} > \underline{l}\). The configuration of the VLP is the vector
\(\mathbf{q} := (q,l) \in \mathbb{S}^1 \times [\underline{l},\overline{l}]\).

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos{q}\dot{q}, \dot{l}\cos(q) - l\sin{q}\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we find the Hamiltonian is equal to the total mechanical energy of the system:
\[
   \mathcal{H} = E = \frac{1}{2}\dot{\mathbf{p}}\tpose \Minv(\mathbf{q})
   \dot{\mathbf{p}}+V(\mathbf{q})
\]
Taking the appropriate derivatives, the dynamics of the VLP in Hamiltonian form
are described in (\ref{eqn:vlp-hamiltonian-with-pl}). 
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q) \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting colinearly with direction of the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, there is no lateral force such as when a person is pushed on a swing.
In this way, the dynamics of \((q,p)\) and \((l,p_l)\) are decoupled.

This decoupling is extremely useful in simplifying the dynamics. Since one can
write set \(\dot{p}_l\) to any desired function, we can suppose (with some abuse
of notation) that \(l\) is tracking some function \(l(t)\). 
The closed loop dynamics of the system will be described exclusively by
\((\dot{q},\dot{p})\) with \(l = l(t)\); 
the fact that \(p_l\) does not appear in these closed-loop dynamics allows us to
ignore the subdynamics \((\dot{l},\dot{p}_l)\) entirely.

What this means is we can treat \(l\) as the control input directly, rather than
modelling it as a configuration variable. Re-deriving the Hamiltonian and the
dynamics with this in mind, we get the system 
(\ref{eqn:vlp-hamiltonian}) with phase \((q,p) \in \mathbb{S}^1 \times \R\). 
Note that the control input \(l(t)\) and its derivative \(\dot{l}(t)\) are both
known variables.
\begin{align}\label{eqn:vlp-hamiltonian}
   \mathcal{H}(q,p) &= \frac{p^2}{ml^2} - \frac{1}{2}\dot{l}^2 - mgl\cos(q) \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{p} = -mgl\sin(q) \\
      \end{cases}\nonumber
\end{align}
In particular, the Hamiltonian of this simplified model is no longer equal to
the total mechanical energy of the system, which is given by
(\ref{eqn:vlp-energy}).
\begin{equation}\label{eqn:vlp-energy}
   E(q,p) = \frac{p^2}{ml^2} + \frac{1}{2}\dot{l}^2 - mgl\cos(q)
\end{equation}
\section{The VLP Constraint}
% TODO: Go through the development of the constraint and prove it gains energy
% TODO: Explain how we came up with the constraint first. Talk about the child
% on a swing, the paper showing the time-optimal control input, and describe why
% we want to remove time from the equation. Discuss the lmin and lmax choices.
% Why do we need to use a VNHC instead of a VHC: we are modelling the
% time-optimal controller in a time-independent way, which requires knowledge of
% which quadrant in (q,p)-space we're in. This means using the angle theta,
% which is inherently a VNHC.
